\documentclass[12pt]{article}
\pagestyle{plain}
\setlength {\topmargin} {-1.0cm}
\setlength {\textwidth} {16.0cm}
\setlength {\textheight} {24cm}
%\voffset -1cm
\pagestyle{empty}
\setlength {\oddsidemargin} {0cm}
\begin{document}

\centerline {\Large {\bf {SSE/AVX and GPU2 with NBODY6}}}
\bigskip
\bigskip
\centerline {\Large {Keigo Nitadori and Sverre Aarseth~~~}}
\bigskip
\medskip
\bigskip
We provide some notes on the use of the SSE (Streaming SIMD Extension),
AVX (Advanced Vector eXtension)
and GPU (Graphics Processing Unit) versions for running
{\tt NBODY6}, first developed at the Institute of Astronomy in April 2008. 
As before, the standard code is compiled by typing {\tt make nbody6} in the
directory {\tt Ncode} and there are no new routines inside the {\tt Makefile}.
For the GPU2 version, type {\tt make gpu} in dir {\tt GPU2}.
\medskip

Hardware requirements for the SSE/AVX/GPU versions are either a multi-core
CPU of the type x86 or x86\_64 processors with SSE/AVX support and/or
a GPU with CUDA support. Core i5/i7 with 64-bit OS is recommended for high
performance calculations with AVX.
GeForce GTX 660Ti or GTX 680 is adequate for a single GPU.
\medskip

Software requirement: 
%GCC officially supports OpenMP (option {\tt -fopenmp}) from
%version 4.2.
%However, CUDA 1.1 does not work with GCC 4.2.
%Hence in some distributions, GCC 4.2 is needed for host compilation and
%4.1 for the GPU code.
%Since Fedora with GCC 4.1 supports OpenMP unoficially, it can be used
%for both compilations.
AVX requires support of both the compiler and kernel.
For simplicity, we mention that combination of CentOS 6.x
and CUDA 4.x or later can compile and execute the code successfully.

\medskip

The directory {\tt GPU2} has several extra Fortran routines which contain the
new procedures, as well as some modified standard routines.
The subdirectory {\tt lib} holds the GPU library for regular force.
To obtain the GPU2 version {\tt nbody6.gpu}, type {\tt make gpu} in the 
directory {\tt GPU2} while {\tt make avx} produces {\tt nbody6.avx}.
In both versions we use the OpenMP directives in some routines.
The executable is sent to the run subdirectory {\tt GPU2/run} which
also contains simple input templates (or see dir {\tt Docs}).
Input for different simulations remains as before, with most options
having the usual meaning.
\medskip

Users can specify the number of threads per process by setting the
environment variable {\tt OMP\_NUM\_THREADS}.
Since we use multiple cores, the CPU time in the output is larger
than the wall-clock time in the ADJUST line. The actual time for data
send and gravity calculation is given on the screen at the end, together
with the corresponding Gflops.

\medskip
Some comments on extra routines in dir GPU2 or {\tt lib}.

\medskip
\noindent
gpunb.velocity.cu: main routine for GPU library in CUDA (dir {\tt lib}).

\medskip
\noindent
intgrt.omp.f: integration flow control for GPU or SSE (also parallelized).

\medskip
\noindent
repair.f: modification of array LISTQ after new or terminated KS.

\medskip
\noindent
jpred.f: standard prediction of X \& XDOT and resolved KS components (J $>$ N).

\medskip
\noindent
jpred2.f: neighbour prediction of single particle (local arrays).

\medskip
\noindent
ksres3.f: transformed coordinates and velocities of KS pairs (local arrays).

\medskip
\noindent
cxvpred.cpp: full X \& XDOT prediction in C++ except for TPRED(J)=TIME.

\medskip
\noindent
gpucor.f: regular force corrector and irregular force loop.

\medskip
\noindent
cmfirr.f: irregular force on perturbed c.m.'s.

\medskip
\noindent
cmfirr2.f: irregular force on singles from c.m.'s at regular force times.

\medskip
\noindent
kspert.f: KS perturbation force loops done in C++ by cnbint.cpp.

\medskip
\noindent
nbintp.f: parallel irregular force corrector with fast neighbour force.

\medskip
\noindent
nbint.f: fast neighbour force ($<$ NPMAX), corrector \& decision-making.

\medskip
\noindent
cnbint.cpp: neighbour force loop (in C++ with SSE).

\medskip
\noindent
adjust.f: standard energy check routine but calls energy2.f.

\medskip
\noindent
gpupot.gpu.cu: fast evaluation of all potentials on GPU (in dir {\tt lib}).

\medskip
\noindent
energy2.f: summation of individual potentials after differential correction.

\medskip
\noindent
phicor.f: differential potential corrections due to binary interactions.

\medskip
\noindent
swap.f: randomized particle swapping at T = 0 (reduces crowding).

\medskip
\medskip
\noindent
Optimization:

\medskip
Optimized performance is achieved by minimizing the number of overflows
which results in the last block members ($<$ NIMAX) being recalculated.
However, small average neighbour numbers (also in \#9 output line) may affect
the accuracy. Based on preliminary tests, a relatively large value of
NNBMAX and option \#40 = 2 (or 3 for decreasing NNBMAX after escape)
appears to be a good strategy.
The fast routine cnbint.cpp is used by gpucor.f, nbint.f, nbintp.f and kspert.f.
Note that all arguments are offset by -1 in cnbint.cpp for consistency with
the C++ convention.
Later we included NBMAX in common6.h and updated it in adjust.f as
MIN(NNBMAX+150,LMAX-5). This is now used as a practical upper limit by
GPUNB\_REGF which reduces overflows (various conditions involving
NNBMAX in gpucor.f have been removed).

\medskip
\medskip
\noindent
%Overflows:
%
%\medskip
%Overflows may occur in two different ways.
%There are 32 blocks for use on the GPU.
%So if the maximum neighbour number is 400 and the indices are
%distributed evenly, there would be 400/32 members in each block.
%Thus a random distribution
%would be within the permitted range nearly all the time.
%However, crowding in the first bin due to mass segregation and
%terminated KS components with small time-steps may occur later.
%The former effect is alleviated by initial random swapping of particle
%indices (but {\it not} names) after data allocation.
%Hence neighbour lists are sequential with masses randomized.

\medskip
\medskip
\noindent
Options:

\medskip
In order to save time on the host (large $N$ only), \#38 = 2 restricts the
neighbour force derivative corrections to $1 \,\%$ regular force change,
while for \#38 = 1 all corrections are done.
The CPU time may also be reduced by saving the common blocks on fort.1
only at every main output as a backup for rare restarts (\#1 = 2 and \#2 = 0).
Option \#40 $>= 2$ stabilizes the average neighbour number (in adjust.f)
on a fraction of the maximum (e.g. NNBMAX/5) for small overflow numbers.
The overflow counters (\#9 OVERFLOWS; current and accumulated) at main
output provide useful diagnostics of the behaviour (\#33 $>$= 2).
To be consistent with decreasing particle numbers, the maximum membership
is reduced by a square root relation scaled by the initial value (\#40 = 3).

\medskip
\medskip
\noindent
GPUIRR

\medskip
The new irregular library deals with prediction of active particles and
evaluation of neighbour forces.
In the case of no regular force calculation,
the predictions are done by GPUIRR\_PRED\_ACT or PRED\_ALL.
Here the parameter NPACT, defined in intgrt.omp.f, is used to decide between
the two versions of prediction, respectively.
Irregular forces for active particles are obtained by the SSE vector
procedure GPUIRR\_FIRR\_VEC.
Finally corrected quantities are sent to this library by GPUIRR\_SET\_JP
at the end of the cycle.

\end{document}
